---
class: "Programming R for Analytics"
title: "Final project: Car crashes in Allegheny county"
author: "Ricardo Ibarra-Gil (ribarrag)"
date: "December 10, 2021"
output: 
  github_document
    #toc: true
    #toc_depth: 3
always_allow_html: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r Libraries and data set, message=FALSE}
# import libraries and data set
library(readr)
library(tidyverse)
library(MASS)
library(corrplot)  
library(plotly)
library(knitr)
library(ggplot2)
crashes <- read_csv("2c13021f-74a9-4289-a1e5-fe0472c89881.csv")
```
### Introduction
This project looks into data about car crashes in Allegheny County. Its main reserach purposes are: 

**1. Provide insightful descriptive statistics about car crashes in Allegheny county.** Present statistics and patterns in the data in a clear and useful way for public policy decisions.  

**2. Run a preliminary model.** Find factors that potentially explain a target variable called 'Severity', that measures the gravity of each car crash, to provide additional statistical analysis.

The ultimate goal of the project is to develop a document that could serve as support for policy decisions regarding road safety. It could assist policy makers to best allocate resources to diminish the negative social impacts of car crashes.  

### 1. About the data set
The original data set was accessed through the Western Pennsylvania Regional Data Center^[The data set can be downloaded from: https://data.wprdc.org/dataset/allegheny-county-crash-data/resource/2c13021f-74a9-4289-a1e5-fe0472c89881] (see footnote for link to data set). 

In its original format and content, the data set has `r nrow(crashes)` rows and `r ncol(crashes)` columns. It comprises all car crashes registered in Allegheny County from 2004 to 2020 (each row represents one accident). The data underwent a cleaning and variable selection process that is detailed in the following section of the document.

The document comprises four sections:  

* **Section 2** includes the cleaning and preprocessing of the data set, which also serves as a first glance into the data and its structure.
* **Section 3** presents the Exploratory Descriptive Analysis that provides more in-depth insights.
* **Section 4** covers a multinomial logistic regression model, to explore which factors contribute to the severity of car crashes.
* Finally, **section 5** includes conclusions and future work.


### 2. Data cleaning  
The variables of interest for this document were selected, and put into 5 subsets for preprocessing: crash consequences, driver conditions, time and place of crash, raod and vehicles variables, and age of drivers:
```{r Variables selection} 
# create subcategories with variables
crash_vars <- c('PROPERTY_DAMAGE_ONLY','INJURY', 'INJURY_COUNT', 'INJURY_OR_FATAL', 'BELTED_DEATH_COUNT', 'FATAL', 'FATAL_COUNT', 'PERSON_COUNT')
driver_conditions <- c('AGGRESSIVE_DRIVING', 'CELL_PHONE', 'DRINKING_DRIVER', 'DISTRACTED', 'FATIGUE_ASLEEP', 'ILLEGAL_DRUG_RELATED', 'RUNNING_RED_LT', 'RUNNING_STOP_SIGN', 'SPEEDING','TAILGATING', 'UNBELTED')
time_place <- c('DAY_OF_WEEK', 'HOUR_OF_DAY','ILLUMINATION_DARK', 'WEATHER')
road_vehicles <- c('CURVED_ROAD', 'ROAD_CONDITION', 'INTERSECTION', 'LOCAL_ROAD', 'URBAN_RURAL', 
                   'SCHOOL_BUS', 'SPEED_LIMIT', 'WET_ROAD', 'VEHICLE_COUNT')
driver_age <- c('DRIVER_16YR', 'DRIVER_17YR', 'DRIVER_18YR', 'DRIVER_19YR', 'DRIVER_20YR', 
                'DRIVER_50_64YR', 'DRIVER_65_74YR', 'DRIVER_75PLUS')
all_vars <- c(driver_conditions, time_place, crash_vars, driver_age, road_vehicles)
# car_crashes will be my main data set
crashes_df = subset(crashes, select = all_vars)
```

**1. Variables related to the results of a crash**. Like number of injured, fatalities, if the crash resulted only in property damage, among others. 

```{r Summary function}
# This is a brief function created to have only mean, min and max from the variables, and ignoring NAs, instead of having all quartiles as with a 'traditional' summary.
summary_functions <- function(i) {c(mean = mean(i,  na.rm = TRUE), min = min(i,  na.rm = TRUE), max = max(i,  na.rm = TRUE), stdv = sd(i, na.rm = TRUE)) } 
```

As basic as they can be, these descriptive statistics show interesting information. 

For instance, it can be seen that, on average, `r round((mean(crashes_df$PROPERTY_DAMAGE_ONLY))*100, 2)`% of car crashes had only property damage, while  `r round((mean(crashes_df$INJURY))*100, 2)`% had one or more people injured, and `r round((mean(crashes_df$FATAL))*100, 2)`% had at least one casualty.

```{r Car crash variables 1}
# This applies the defined function to only the subset of variables related to car crash results
kable(round(sapply(subset(crashes_df, select = crash_vars[1:4]), summary_functions), 3), results = 'asis')
```

The standard deviation for INJURY_COUNT tells us that, although there are some observations with high values (like the max = `r max(crashes_df$INJURY_COUNT)`), the vast majority of the data for that variable is concentrated around its mean (`r round(mean(crashes_df$INJURY_COUNT), 3)` injured per car crash).

As for the min and max of these variables, since most of them are dummies, they should be 0 and 1, except for the variables *count*, that show counts of people in each accident. 

```{r Car crash variables 2}
# This applies the defined function to only the subset of variables related to car crash results
kable(round(sapply(subset(crashes_df, select = crash_vars[5:8]), summary_functions), 3), results = 'asis')
```

For example, PERSON_COUNT is how many people were involved in the accident. It has a standard deviation of `r round(sd(crashes_df$PERSON_COUNT), 3)`, which indicates that most observations are concentrated around its average. Its maximum value of 99 is already a redflag in our data set: it is not very probable that there was a crash involving 99 people, so this can be an outlier that can even be an encoding error. To further examine this data, its boxplot is graphed:

```{r Boxplot Person Count, fig.dim = c(7, 5), fig.cap = 'Boxplot. Count of persons involved in crash', fig.align = 'center'}
# create a box plotw with ggplot 
ggplot(crashes_df, aes(y=PERSON_COUNT)) + geom_boxplot(color = 'steelblue') + 
  labs(title = 'Graph 1. Box plot for count of people involved in accident', y = 'Person Count') +
  theme(plot.title = element_text(hjust = 0.5, # I will use Helvetica bold  for titles of graphs and centered titles
                                  size = 16, 
                                  family = 'Helvetica-Bold'), 
        axis.title.y = element_text(size = 14, 
                                    margin = margin(0, 5, 0, 0)),
        plot.margin = margin(20, 20, 15, 15),
        panel.background = element_rect(fill = 'white'), # set the back of the graph to white
        panel.grid.major.y = element_line(size = .08, # leave only thin dashed lines
                                          linetype = 'dashed', 
                                          color = 'black'))
crashes_df$PERSON_COUNT[crashes_df$PERSON_COUNT > 20] = NA # this is setting values above 20 to missing values
```

The boxplot in **Graph 1** helps define the presence of outliers. In this case, a decision about our data had to be made: outliers in the variable PERSON_COUNT will be considered those crashes in which more than 20 individuals were involved, and in this case were set to NA. As mentioned, some can even be data entry errors.

As can be seen in the table below, all variables in this subset are numerical. Most of them are dummies (indicate the presence/absence of an attribute), and some are ordinal variables (count the number of times any attribute was present in a car crash).

```{r Subset crash vars structure}
str(subset(crashes_df, select = crash_vars))  #asks for the structure of the variables in the category crash_vars
```

From these variables, a **'SEVERITY'** variable was created to represent how bad the crash was in terms of damages and injured people. The severity variable is classified according to the following criteria:

* 0 = No property damage, no injured, no death  
* 1 = Only property damage  
* 2 = One injured, no death  
* 3 = Several injured, no death  
* 4 = Death, can be one or multiple  

This variable will serve as dependant varable for the regression analysis of section 4.


```{r Create variable severity}
# create the severity variable according to the description in the text
crashes_df <- crashes_df %>% 
  mutate(SEVERITY = if_else((PROPERTY_DAMAGE_ONLY == 0) & (INJURY == 0) & (INJURY_COUNT == 0) & (FATAL == 0), 0,
        + ifelse ((PROPERTY_DAMAGE_ONLY == 1) & (INJURY == 0) & (INJURY_COUNT == 0), 1,
        + ifelse ((INJURY == 1) & (INJURY_COUNT == 1) & (FATAL == 0) & (FATAL_COUNT == 0), 2, 
        + ifelse ((INJURY == 1) & (INJURY_COUNT > 1)  & (FATAL == 0) & (FATAL_COUNT == 0), 3,
        + ifelse ((FATAL == 1), 4, NA))))))
crashes_df$SEVERITY_L <- factor(crashes_df$SEVERITY,
                                 levels = c(0, 1, 2, 3, 4),
                                 labels = c("(0) No Damages", "(1) Only property damages", "(2) Only 1 injured", '(3) Several injured, no deaths', '(4) Fatalities (1 or many)'))
```

The mean of SEVERITY is `r round(mean(crashes_df$SEVERITY, na.rm = TRUE),3)`. This indicates that the consequence of the average car crash in Allegheny county from 2004 to 2020 lies between only property damage (1), and 1 person injured with no fatalities (2).

However, it is also interesting to the see the numbers in absolute terms:

```{r Table Severity variable}
# display the categories of var SEVERITY with its frequency
kable(table(crashes_df$SEVERITY_L), col.names = c('Severity of crash', 'Frequency'))
```

In the table above it can fully appreciated that a very high number of crashes result in only property damage (`r table(crashes_df$SEVERITY)[2]` out of `r sum(table(crashes_df$SEVERITY))`).
Also, that there were `r table(crashes_df$SEVERITY)[4]` crashes that resulted in several people injured, and `r table(crashes_df$SEVERITY)[5]` that resulted in one or more fatalities.

**2. Variables related to the drivers' conditions**. The following variables are all dummies (0,1), so it is more interesting to show the mean for each of them; but also the min and max to verify for any potential missing values encoded as numeric:
```{r Table Driver conditions}
# I separated the summary of the variables in driving conditions, so they appear in two different rows, and not all 11 in one single row, because it conflicts with margins
kable(round(sapply(subset(crashes_df, select = c(driver_conditions[1:5])), summary_functions), 3))
kable(round(sapply(subset(crashes_df, select = c(driver_conditions[6:11])), summary_functions), 3))
```

It can be seen that there are no variables with maximum greater than 1 (which is correct, since all are dummies). It is interesting that in a bit more than half of the accidents, there was some AGGRESSIVE_DRIVING behavior; in `r round((mean(crashes_df$DISTRACTED))*100, 2)`% there was at least one distracted driver, and in `r round((mean(crashes_df$UNBELTED))*100, 2)`% of all crashes, at least one of the drivers involved in the accident was not wearing a set belt.

As can be seen below, all variables are numeric.
```{r Subset driver conditions var structure}
# print structure of variables in the driver_conditions subset
str(subset(crashes_df, select = c(driver_conditions)))
```
  
The dataset has dummy variables for certain age cohorts, but unfortunately not for all. 

```{r Driver age variables}
#same separation of variables descriptors to fit the margins
kable(round(sapply(subset(crashes_df, select = c(driver_age[1:4])), summary_functions),3))
kable(round(sapply(subset(crashes_df, select = c(driver_age[5:8])), summary_functions),3))
```

From a first glance at the data we can see that there is an anomaly with the dummy DRIVER_16YR (it indicates if there was (or not) at least one driver of age 16 in the car crash). This dummy
should not have values greater than 1. So, the values that are greater than 1 (which are exactly: `r length(crashes_df$DRIVER_16YR[crashes_df$DRIVER_16YR == 2])`, and both set to 2) will be set to missing values.

As can be seen below. All these variables have only numeric values only:

```{r Driver age var structure}
# set observations that are 2 to NA because variable is dummy (0,1 only)
crashes_df$DRIVER_16YR[crashes_df$DRIVER_16YR == 2]= NA
str(subset(crashes_df, select = c(driver_age)))
```


**3. Time and place of the accidents.**  This category contains only a few variables
The table below indicates that one of them (HOUR_OF_DAY) is of type `r typeof(crashes_df$HOUR_OF_DAY)`.

```{r Time place variables}
# print structure of variables in the time_place subset
str(subset(crashes_df, select = c(time_place)))
```
It stands out that the WEATHER variable has a high max:
```{r Numeric Time place variables summary}
kable(round(sapply(subset(crashes_df, select = c('DAY_OF_WEEK', 'ILLUMINATION_DARK', 'WEATHER')), summary_functions), 3))
```
This high max value of WEATHER should be explored in more detail. These are the values taken by this variable:
```{r Weather var table}
# show the unique entries of weather
unique(crashes_df$WEATHER)
```
The NAs will be dropped at the end of this section, but there are some 98's and 99's, and according to the documentation, there are only 10 valid values. So I will also set the observations above 10 as missing values (there are not many, only `r round((table(crashes_df$WEATHER)[11] + table(crashes_df$WEATHER)[12]) / sum(table(crashes_df$WEATHER)) * 10,3)`% of total values are 98 or 99).
```{r}
# set weather variable observations greater than 10 to NA
crashes_df$WEATHER[crashes_df$WEATHER > 10] = NA
```

And the variable is set into factor, and labels are defined for each level.

```{r Weather variable cleaning}
crashes_df$WEATHER <- factor(crashes_df$WEATHER,
                              levels = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10),
                              labels = c("Blowing sand / soil", "Blowing snow", "Clear", 'Cloudy', 'Fog/Smoke', 'Freezing rain', 'Rain', 'Severe crosswinds', 'Sleet / hail', 'Snow'))
```

Something similar is happening with the variable HOUR_OF_DAY. From documentation (and the fact that there are only 24 hours in a day) we know that it can only go from 00 to 23. However, the variable has some 99s in it, which will be set to missing values:

```{r}
unique(crashes_df$HOUR_OF_DAY)
```

```{r Hour of day variable}
crashes_df$HOUR_OF_DAY[crashes_df$HOUR_OF_DAY == 99] = NA
```

And it will also be set to factor, and set the label for its levels.
```{r Day of week var cleaning}
crashes_df$DAY_OF_WEEK <- factor(crashes_df$DAY_OF_WEEK,
                                   levels = c(1, 2, 3, 4, 5, 6, 7),
                                   labels = c("Sunday", "Monday", "Tuesday", 'Wednesday', 'Thursday', 'Friday', 'Saturday'))
```

**4. Variables about rod and vehicles.** The final of the data categories.
```{r Road vehicles varables}
str(subset(crashes_df, select = c(road_vehicles)))
```

All of the variables are numeric, and their summary statistics are:
```{r}
kable(round(sapply(subset(crashes_df, select = c(road_vehicles)[1:5]), summary_functions), 3))
kable(round(sapply(subset(crashes_df, select = c(road_vehicles)[6:9]), summary_functions), 3))
```

We can see that ROAD_CONDITION has max values of 98 and 99, which doesn't make much sense
```{r Road Condition cleaning}
unique(crashes_df$ROAD_CONDITION)
```

After checking the data set documentation,98 and 99 from ROAD_CONDITION are reclassified as NA, and 8, 9 and 22 into 'Other'.

```{r}
crashes_df$ROAD_CONDITION[crashes_df$ROAD_CONDITION > 22] = NA
crashes_df$ROAD_CONDITION <- factor(crashes_df$ROAD_CONDITION,
                                      levels = c(0, 1, 2, 3, 4, 5, 6, 7, c(8, 9, 22)),
                                      labels = c("Dry", "Wet", "Sand / mud / gravel", 'Snow', 'Slush', 'Ice', 'Ice patches', 'Water', 'Other', 'Other', 'Other'))
```

Now that we have taken a glance at the data, and made all changes to clean it, the missing values are dropped, and the data is ready to start analyzing it in more depth:
```{r}
crashes_df <- crashes_df %>% drop_na()
```

I have a data set clean and ready to work with.
Also, now I have a pretty good idea of what is in my data set

### 3. Exploratory Descriptive Analysis

A first interesting insight about car crashes in Allegheny County is how they are distributed in time.

```{r Graph Hours, fig.align = 'center', fig.cap = 'Car crashes per hour of day', fig.dim = c(6,6) }
ggplot(crashes_df, aes(x=(HOUR_OF_DAY)))+
  geom_bar(stat="count", width=0.7, fill="steelblue") +
  labs(title = 'Graph 2. Car crashes per hour of the day', y = 'Crashes count', x = 'Hour of the day') + 
  scale_y_continuous(breaks=seq(0,14000,2000)) +
  theme(plot.title = element_text(hjust = 0.5, 
                                size = 16, 
                                family = 'Helvetica-Bold'),
      axis.text = element_text(size = 10), 
      axis.title.x = element_text(size = 14, 
                                  margin = margin(5, 0, 0, 0)),
      axis.title.y = element_text(size = 14, 
                                  margin = margin(0, 5, 0, 0)),
      axis.text.x = element_text(margin = margin(0, 0, 0, 0)),
      axis.ticks = element_blank(),
      plot.margin = margin(20, 20, 15, 15), 
      panel.background = element_rect(fill = 'white'),
      panel.grid.major.y = element_line(size = .08, 
                                        linetype = 'dashed', 
                                        color = 'black'))
```

It can be seen in **Graph 2** that, in absolute terms, most crashes happen in the late afternoon-early evening, from 3:00 to 5:00 PM., with a peak at 17:00 hours. This is interesting and a bit surprising. 
 
It cluld be expected that more accidents happened towards 19:00 when (I believe) people leave their jobs and go home. A possible explanation could be that before peak hour, people try to get home quickly before traffic starts, and drive with less care, so there is an increased number of accidents around 5:00 PM.


**Graph 3** shows the relative distribution of crashes with fatal outcomes (one or more people died in the accident).
```{r Avg crashes with fatal by hour, fig.dim = c(8, 6), fig.align = 'center'}
barplot(tapply(crashes_df$FATAL, list(crashes_df$HOUR_OF_DAY), mean), main = 'Graph 3. Average crashes with fatal outcomes per hour of the day', xlab = 'Hour of the day', ylab = 'Avg crashes with fatalities', col = 'steelblue', las = 2, ylim=c(0,.02))
```

It should be noticed from **Graph 3** that the y axis has a range of low numbers because the number of crashes with casualties is relatively small (as seen in Section 2) as compared to the total number of car accidents. Nevertheless, we can still observe that crashes with deaths concentrate around the first hours of the day (00:00 and 5:00 am) and have a peak at 02:00 am. 

```{r Crashes by day of week, fig.dim = c(7, 6), fig.align = 'center'}
ggplot(crashes_df, aes(x=(DAY_OF_WEEK)))+
  geom_bar(stat="count", width=0.7, fill = "steelblue") +
  labs(title = 'Graph 4. Car crashes per day of the week', y = 'Crashes count', x = 'Day of the week') + 
  scale_y_continuous(breaks=seq(20000,32000,2000)) +
  coord_cartesian(ylim = c(20000, 32000)) +
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 16, 
                                  family = 'Helvetica-Bold'),
        axis.text = element_text(size = 10), 
        axis.title.x = element_text(size = 14, 
                                    margin = margin(10, 0, 0, 0)),
        axis.title.y = element_text(size = 14, 
                                    margin = margin(0, 5, 0, 0)),
        axis.text.x = element_text(margin = margin(10, 0, 0, 0)),
        axis.ticks = element_blank(),
        plot.margin = margin(20, 20, 15, 15), 
        panel.background = element_rect(fill = 'white'),
        panel.grid.major.y = element_line(size = .08, 
                                          linetype = 'dashed', 
                                          color = 'black'))
```

When we turn our analysis to day of the week, it can be observed (**Graph 4**) that Sunday is the day when less crashes happen (11.90% of total crashes), which is perhaps associated with the fact that a lot of people rest on that day, and don't use their cars. Then crashes go up during the week and reach a peak on Fridays (16.74% of total crashes).


```{r Plotly graph Day Time, fig.dim = c(8,10), fig.align = 'center'}
# create a proportions table for the graph
data_plotly <- prop.table(table(crashes_df$DAY_OF_WEEK, crashes_df$HOUR_OF_DAY))
# Here, i convert the column names (hours) into format 00:00, because otherwise they are presented like numbers (0, 1, 2, ...) and it is not clear that these are hours. 
colnames(data_plotly) <- format(strptime(as.numeric(colnames(data_plotly)), format = "%H"), format = "%H:%M")
# Transpose the matrix because im interested in changing the column names
data_plotly <- t(as.matrix(data_plotly))
# change column names to name of day
colnames(data_plotly)[1] <- 'SUN'
colnames(data_plotly)[2] <- 'MON'
colnames(data_plotly)[3] <- 'TUE'
colnames(data_plotly)[4] <- 'WED'
colnames(data_plotly)[5] <- 'THU'
colnames(data_plotly)[6] <- 'FRI'
colnames(data_plotly)[7] <- 'SAT'
data_plotly <- round(data_plotly*100,2)
# finally graph that information
p <- plot_ly(x = colnames(data_plotly), y = rownames(data_plotly), 
             z = data_plotly, 
             type = "heatmap", 
             colorscale= "RdBu", # set colors to from red to blue
             showscale = F) %>%
  layout(title = list(text ='Graph 5. Car crashes per time and day (% of car crashes)', font = list(family = "Helvetica-Bold", size = 16, color = 'black')), # had to change the order to show hours of day in increasing order
         yaxis = list(dtick = 1, ticks = '', ticklabelposition = "inside", autorange = 'reversed'), 
         xaxis = list(side = 'top', tickangle = 0),
         margin = list(l = 40, r = 40, b = 40, t = 80))
p
```

An analysis of both day of the week and time of the day is displayed in **Graph 5**. It is an interactive graph, so the user can hover over it, and it will display the Day, Time and Percentage of crashes. The color of the square indicates in what day and time when most crashes occur (blue for less occurrences and red for more). For instance, on a Friday at 17:00 hrs., 1.29% of all crashes of Allegheny County take place.

Moving towards other interesting variables, **Graph 6** displays the percentage of crashes that had fatalities, by type of weather. 

```{r Crashes with fatalities by weather, fig.dim = c(8, 6), fig.align = 'center'}
ggplot(crashes_df, aes(x = WEATHER, y = FATAL*100)) + 
  geom_bar(stat = "summary", fun = "mean",  width=0.7, fill="steelblue") + 
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = 'Graph 6. Percentage of crashes with fatalities per type of weather', y = 'Percentage', x = 'Weather') + 
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 16, 
                                  family = 'Helvetica-Bold', 
                                  margin = margin(0, 0, 15, 0)),
        axis.text = element_text(size = 10), 
        axis.title.x = element_text(size = 14, 
                                    margin = margin(5, 0, 0, 0)),
        axis.title.y = element_text(size = 14, 
                                    margin = margin(0, 5, 0, 0)),
        axis.text.x = element_text(margin = margin(0, 0, 0, 0)),
        axis.ticks = element_blank(),
        plot.margin = margin(20, 20, 15, 15), 
        panel.background = element_rect(fill = 'white'),
        panel.grid.major.y = element_line(size = .08, 
                                          linetype = 'dashed', 
                                          color = 'black'))
```

In almost `r round((tapply(crashes_df$FATAL, list(crashes_df$WEATHER), mean)[9])*100,2)`% of crashes where there was sleet or hail, there was at least one fatality. This  compares, for example, with around 0.5% of crashes with fatalities out of total crashes when there is sand or soil, snow, when the sky is clear, or when there are severe crosswinds. This should certainly be of interest for authorities to invest more in training people to drive during this type of weather in order to reduce car crash fatalities.

**Graph 7** shows the average severity of crashes by road condition, separating between speeding and not speeding. Let's remeber that severity is a variable that goes from 0 to 4 and measures the negative impact of a car crash, so the higher the mean, the most negative the consequence of a crash. 

```{r Severity by Road Condition and Speeding, fig.dim = c(8, 6), fig.align = 'center'}
ggplot(crashes_df, aes(x = ROAD_CONDITION, y = SEVERITY)) + geom_bar(stat = "summary", fun = "mean") + facet_grid(crashes_df$SPEEDING) + # here, the facet dividing over speeding
  scale_y_continuous(breaks=seq(1,2,0.4)) + coord_cartesian(ylim = c(1.2, 1.8)) +
  geom_bar(stat = "summary", fun = "mean", fill="steelblue") +
  labs(title = 'Graph 7. Avg Severity of crashes per Road Condition and Speeding', y = 'Avg. Severity', x = 'Road Condition') + 
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 16, 
                                  family = 'Helvetica-Bold', 
                                  margin = margin(0, 0, 15, 0)),
        axis.text = element_text(size = 10), 
        axis.title.x = element_text(size = 14, 
                                    margin = margin(5, 0, 0, 0)),
        axis.title.y = element_text(size = 14, 
                                    margin = margin(0, 5, 0, 0)),
        axis.text.x = element_text(margin = margin(0, 0, 0, 0), angle = 90),
        axis.ticks = element_blank(),
        plot.margin = margin(20, 20, 15, 15), 
        panel.background = element_rect(fill = 'white'),
        panel.grid.major.y = element_line(size = .08, 
                                          linetype = 'dashed', 
                                          color = 'black'))
```

The lower part of **Graph 7** is when there was speeding in the accident, and it can be noticed that severity of crashes increases realtively to the upper part of the graph (no speeding) for almost all road conditions, but especially for snow and slush. This could also be of interest for policymakers: targeting speeding might have an impact on reducing severity of car crashes in Allegheny county. 

```{r Definition of ages vars}
ages <- sapply(subset(crashes_df, select = c(driver_age)), sum)
age <- as.data.frame(ages)
```

The data set had dummy variables for certain age cohorts. Unfortunately, there is not a variable with all driver's age, but only for those whose age fall in the cohorts displayed in the following graph:

```{r Crashes by age, fig.dim = c(8, 6), fig.align = 'center'}
ggplot(age, aes(x = rownames(age), ages)) + geom_bar(stat='identity', width=0.7, fill="steelblue") +
  labs(title = 'Graph 8. Crashes by age', 
       subtitle = '(available age cohorts only)', 
       x = 'Age cohorts', 
       y = 'Crashes count') +
   ylim(0, 60000) +
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 16, 
                                  family = 'Helvetica-Bold'),
        plot.subtitle = element_text(hjust = 0.5, 
                                     size = 12, 
                                     family = 'Helvetica', 
                                     margin = margin(0, 0, 5, 0)),
        axis.text = element_text(size = 10), 
        axis.title.x = element_text(size = 14, 
                                    margin = margin(5, 0, 0, 0)),
        axis.title.y = element_text(size = 14, 
                                    margin = margin(0, 5, 0, 0)),
        axis.text.x = element_text(margin = margin(0, 0, 0, 0)),
        axis.ticks = element_blank(),
        plot.margin = margin(20, 20, 15, 15), 
        panel.background = element_rect(fill = 'white'),
        panel.grid.major.y = element_line(size = .08, 
                                          linetype = 'dashed', 
                                          color = 'black')) +
  scale_x_discrete(labels = c('DRIVER_16YR' = '16','DRIVER_17YR' = '17',
                              'DRIVER_18YR' = '18','DRIVER_19YR' = '19',
                              'DRIVER_20YR' = '20','DRIVER_50_64YR' = '50-64',
                              'DRIVER_65_74YR' = '65-75','DRIVER_75PLUS' = '75'))

```

Regarding age, it would be very interesting to analyze if being a young driver is related to more severity of car crashes, compared to other age cohorts in the regression model. So the variable YOUNG_DRIVER is generated, and it is set to 1 if the driver is 16, 17, 18, 19 or 20, and set to 0 otherwise.
```{r Creat Young_Driver var}
crashes_df <- crashes_df %>% 
  mutate(YOUNG_DRIVER = if_else((DRIVER_16YR == 1) | (DRIVER_17YR == 1) | (DRIVER_18YR == 1) | (DRIVER_19YR == 1) | (DRIVER_20YR == 1), 1, 0))
```


Once we have this variable of YOUNG_DRIVERS, it is interesting to analyze if, among young drivers there are some who are more prone to accidents. 

```{r Create input for graph 9}
# only took the first five ages (the young drivers from 'ages')
young_pct <- round((ages[1:5]) *100 / sum(ages[1:5]),2)
young_pct <- as.data.frame(young_pct)
young_pct$`young_cohort` <- rownames(young_pct)
# added a column named prop, to get the folowing graph to work
young_pct <- young_pct %>% 
  mutate(Prop = "Proportions")
```
```{r Young drivers proportions graph, fig.dim = c(7, 5), fig.align = 'center'}
ggplot(young_pct, aes(x = Prop, y = young_cohort, fill = young_cohort)) +
  geom_col()+
  geom_text(aes(label = young_pct),
            position = position_stack(vjust = 0.5)) + coord_flip() + 
  scale_fill_manual(values=c('#3C6EC2', '#4E84C4', '#56B4E9', '#52854C', '#C3D7A4')) +
  labs(title = 'Graph 9. Young driver crashes: percentage of each age cohort', 
       x = '', 
       y = '') +
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 16, 
                                  family = 'Helvetica-Bold', 
                                  margin = margin(10, 0, -40, 0)), 
        panel.background = element_rect(fill = 'white'), legend.position = 'bottom',
        legend.title = element_blank(), legend.text=element_text(size = 10),
        legend.box.margin = margin(-80, 0, 0, 0),
        axis.text = element_blank(), 
        axis.ticks = element_blank()) +
  guides(fill = guide_legend(reverse = TRUE))
```
As **Graph 9 shows**, of all **young** drivers (100%) who were involved in a crash, 25.8% were aged 20 years old, while only 7.4% were 16 years old. This can be partially explained by the fact that there might be more 20 year old drivers than 15 year old drivers, but it is still an interesting graph.



Also about young drivers, we can observe in **Graph 10** the average severity of accidents (represented in the bars) by weather and separated in two grids. The lower one for young drivers, and the upper one for non-young drivers. 

```{r Severity by weather and young driver, fig.dim = c(8, 6), fig.align = 'center'}
ggplot(crashes_df, aes(x = WEATHER, y = SEVERITY)) + geom_bar(stat = "summary", fun = "mean") + facet_grid(crashes_df$YOUNG_DRIVER) + # the facet is dividing over youn_driver variable here
  scale_y_continuous(breaks=seq(1,2,0.4)) + coord_cartesian(ylim = c(1.2, 1.8)) +
  geom_bar(stat = "summary", fun = "mean", fill="steelblue") +
  labs(title = 'Graph 10. Avg Severity of crashes per weather and Young Driver', y = 'Avg. Severity', x = 'Road Condition') + 
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 16, 
                                  family = 'Helvetica-Bold', 
                                  margin = margin(0, 0, 15, 0)),
        axis.text = element_text(size = 10), 
        axis.title.x = element_text(size = 14, 
                                    margin = margin(5, 0, 0, 0)),
        axis.title.y = element_text(size = 14, 
                                    margin = margin(0, 5, 0, 0)),
        axis.text.x = element_text(margin = margin(0, 0, 0, 0), angle = 90),
        axis.ticks = element_blank(),
        plot.margin = margin(20, 20, 15, 15), 
        panel.background = element_rect(fill = 'white'),
        panel.grid.major.y = element_line(size = .08, 
                                          linetype = 'dashed', 
                                          color = 'black'))
```

From **Graph 10** we can say that the severity of accidents is higher among young drivers (lower grid), particularly when there is freezing rain, severe crosswinds and sleet / hail, compared to non-young drivers (upper grid). Maybe a policy to educate young drivers to drive better under those weather conditions could have a meaningful positive impact.

We have analyzed time and weather for crashes and fatalities. For this last part of the Exploratory Data Analysis, there are two graphs about drivers' profile and behavior.

```{r Create vars for graphs 11, 12}
# I create the input for the following graph, needed tables for the subset driver conditions (their means as percentage)
crash_driver <- round((sapply(subset(crashes_df, select = c(driver_conditions)), mean)*100),1)
crash_driver_type <- as.data.frame(crash_driver)
crash_driver_type$`driver type` <- rownames(crash_driver_type)
fatalities <- subset(crashes_df, crashes_df$FATAL == 1)
fatal_driver <- round((sapply(subset(fatalities, select = c(driver_conditions)), mean)*100),1)
fatal_driver_type <- as.data.frame(fatal_driver)
fatal_driver_type$`driver type` <- rownames(fatal_driver_type)
```

```{r Crashes by driver type, fig.align = 'center'}
ggplot(crash_driver_type, aes(x = reorder(`driver type`, crash_driver), y = crash_driver, label = crash_driver)) + 
  geom_point(stat='identity', color="steelblue", size = 12) +
  geom_segment(aes(y = 0, 
                   x = `driver type`, 
                   yend = crash_driver, 
                   xend = `driver type`), 
               color = "steelblue") +
  geom_text(color="white", size = 3.6) + # set color and size for text inside circles
  labs(title="Graph 11. Crashes by driver type", 
       subtitle="(% of crashes with at least one of these drivers)",
       x = 'Type of driver',
       y = '% of crashes') + 
  ylim(0, 60) + # set limits to y axis
  coord_flip() + # revert the fgraph so it appears from left to right instead of from down to up
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 16, 
                                  family = 'Helvetica-Bold'),
        plot.subtitle = element_text(hjust = 0.5, 
                                     size = 12, 
                                     family = 'Helvetica', 
                                     margin = margin(0, 0, 5, 0)),
        axis.text = element_text(size = 10), 
        axis.title.x = element_text(size = 14, 
                                    margin = margin(5, 0, 0, 0)),
        axis.title.y = element_text(size = 14, 
                                    margin = margin(0, 5, 0, 0)),
        axis.text.x = element_text(margin = margin(0, 0, 0, 0)),
        axis.ticks = element_blank(),
        plot.margin = margin(20, 20, 15, 15), 
        panel.background = element_rect(fill = 'white'),
        panel.grid.major.x = element_line(size = .08, # for almost all graphs I set the background to white 
                                          linetype = 'dashed', # but kept some thins dashed lines
                                          color = 'black')) +
        scale_x_discrete(labels = c('AGGRESSIVE_DRIVING' = 'AGGRESSIVE','DRINKING_DRIVER' = 'DRINKING',
                                    'RUNNING_RED_LT' = 'RUNNING RED LIGHT','RUNNING_STOP_SIGN' = 'RUNNING STOP SIGN',
                                    'FATIGUE_ASLEEP' = 'TIRED','CELL_PHONE' = 'DISTRACTED ON PHONE',
                                    'ILLEGAL_DRUG_RELATED' = 'ON ILLEGAL DRUG')) 
```
**Graph 11** shows that 51.32% of crashes had at leat one aggressive driver, by far the most common occurrence. This could hep policy makers to implement measures to decrease stress and anger in the road. Second comes distracted drivers. 

```{r Crashes with fatality by driver type, fig.align = 'center'}

ggplot(fatal_driver_type, aes(x=reorder(`driver type`, fatal_driver), y = fatal_driver, label = fatal_driver)) + 
  geom_point(stat='identity', 
             color = "steelblue", 
             size = 12) +
  geom_segment(aes(y = 0, 
                   x = `driver type`, 
                   yend = fatal_driver, 
                   xend = `driver type`), 
               color = "steelblue") +
  geom_text(color="white", size = 3.6) +
  labs(title = 'Graph 12. Crashes with fatalities by driver type', 
       subtitle = '(% of crashes with fatalities with at least one of these drivers)', 
       x = 'Car crashes with fatalities', 
       y = '% of crashes with fatalities') + 
  ylim(0, 60) +
  coord_flip() +
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 16, 
                                  family = 'Helvetica-Bold'),
        plot.subtitle = element_text(hjust = 0.5, 
                                     size = 12, 
                                     family = 'Helvetica', 
                                     margin = margin(0, 0, 5, 0)),
        axis.text = element_text(size = 10), 
        axis.title.x = element_text(size = 14, 
                                    margin = margin(5, 0, 0, 0)),
        axis.title.y = element_text(size = 14, 
                                    margin = margin(0, 5, 0, 0)),
        axis.text.x = element_text(margin = margin(0, 0, 0, 0)),
        axis.ticks = element_blank(),
        plot.margin = margin(20, 30, 15, 15), 
        panel.background = element_rect(fill = 'white'),
        panel.grid.major.x = element_line(size = .08, 
                                          linetype = 'dashed', 
                                          color = 'black')) +
  scale_x_discrete(labels = c('AGGRESSIVE_DRIVING' = 'AGGRESSIVE','DRINKING_DRIVER' = 'DRINKING',
                              'RUNNING_RED_LT' = 'RUNNING RED LIGHT','RUNNING_STOP_SIGN' = 'RUNNING STOP SIGN',
                              'FATIGUE_ASLEEP' = 'TIRED','CELL_PHONE' = 'DISTRACTED ON PHONE',
                              'ILLEGAL_DRUG_RELATED' = 'ON ILLEGAL DRUG')) 
```
When we restrict only to those accidents with at least one fatality, as in **Graph 12**, again in 51.54% of crashes there was at least one aggressive driver, but then unbelted drivers come second. It means that in 41.48% crashes that resulted in one or more deaths, there were unbelted drivers. Wearing the belt while driving should also be a priority in policies to diminish death on the road, as well as disuading people from drink and driving, and speeding. 

Please notice that the number of graphs 11 and 12 do not add up to 100 becuase there are accidents were the drivers had two or more of the mentioned conditions.


### 4. Regression Analysis
In this last section of the analysis, a couple of correlation matrices are presented. There are some interesting correlations, such as the negative one between aggressive driving and distracted driving. The use of cellphone is related positively with distracted driving, as would be expected. And also tailgating and speeding have a positive relation with aggressive driving, which is also very understandable.

```{r Correlation 1, fig.align = 'center'}
model_X_1 <- c('SEVERITY', 'DRINKING_DRIVER', 'DISTRACTED', 'UNBELTED', 'AGGRESSIVE_DRIVING', 'CELL_PHONE', 'FATIGUE_ASLEEP','ILLEGAL_DRUG_RELATED', 'SPEEDING', 'TAILGATING', 'YOUNG_DRIVER')
# after selecting a subset for correlations, compute correlation
model_X1_cor <- subset(crashes_df, select = model_X_1 )
# show in a graph
corrplot(cor(model_X1_cor))

```
The negative correlation between local road and speed limit was expected: local roads tend to have lower speed limits.

```{r Correlation 2, fig.align = 'center'}
model_X_2 <- c('SEVERITY', 'ILLUMINATION_DARK', 'CURVED_ROAD', 'INTERSECTION', 'SPEED_LIMIT', 'LOCAL_ROAD', 'URBAN_RURAL')
model_X2_cor <- subset(crashes_df, select = model_X_2 )
corrplot(cor(model_X2_cor))
```

For the regression analysis, a multinomial logistic regression is run. In order not to plug everything into the model, I selected the variables that might be more powerful to explain the severity of crashes. Also, some variables were not included for potential multicollinearity, as is the case of the ROAD_CONDITION variable that should suffer from almost perfect correlation with the variable WEATHER.

The target variable is Severity, and the results are presented below:

```{r Regression model}
# specification of regression model
model_1 = polr(SEVERITY_L ~ DRINKING_DRIVER + DISTRACTED + UNBELTED + AGGRESSIVE_DRIVING + CELL_PHONE + FATIGUE_ASLEEP + ILLEGAL_DRUG_RELATED + SPEEDING + TAILGATING + UNBELTED + YOUNG_DRIVER + ILLUMINATION_DARK +  CURVED_ROAD + INTERSECTION + URBAN_RURAL + SPEED_LIMIT + LOCAL_ROAD + as.factor(DAY_OF_WEEK) + as.factor(WEATHER), data = crashes_df, Hess = TRUE)
summary(model_1)
```

I will not describe all the regression ouptut, but rather focus on the most interesting results:

* **Drinking driver.** It has a positive coefficient, meaning that drinking drivers are related to a higher probability of having a crash with more severe consequences. This makes sense, and we have been observing that in the data. One important recommendation for policy makers is to continue focusing on drink and driving.
* **Unbelted.** Unbelted drivers are associated with higher severity of accidents. A policy that enforces wearing the seatbelt could also lead to positive results.
* **Intersection.** Its coefficient has a potivie sign as well, so paying attention to interesections is important. Making sure that signs are visibile, and maybe event randomly patrolling the intersections where most accidents happen coul be an interesting policy.
* **Local road.** Local roads has a negative coefficient, this might be related to speed limits.
* **Snow.** Interestingly, snow has also a negative coefficient. Probably because people are bein more cuatios when driving during snow, and there fore accidents are less severe.

This was an initial regression analysis, and some improvements can be made to have a more robust regression model. Nevertheless, the results are interesting and could be of utility for orienting policy recommendation.

### 5. Conclusions and future work

As a very brief conclusion, the data is very complete and interesting and there are other variables that due to time restrictions were not analysed thouroughly in this project, but that could serve for additional research. 

The results of the descriptive analysis and the regression model are consistent, and give a pretty clear idea of the most important factors that increase or decrease the severity of crashes in Allegheny County. 

Some of the limitations of this analysis include: most variables are catorgircal, so the diversity of graphs to present the data was a little limited; also, it would have been interesting to have a time indicator to perform time series analysis.

### References
I used some references, mainly as help for polishing my graphs.  
https://dplyr.tidyverse.org/reference/summarise.html  
https://stackoverflow.com/questions/24027605/determine-the-number-of-na-values-in-a-column  
Got some help for the use of strptime:  
https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/strptime  
Help with a couple of graphs:  
http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html   
Got some help for graphing with plotly, especially for changing the layout of the graph  
https://stackoverflow.com/questions/37809906/control-which-tick-marks-labels-appear-on-x-axis-in-plotly  
To create a new column based on criteria of other columns:  
https://stackoverflow.com/questions/12125980/how-to-create-a-new-variable-in-a-data-frame-based-on-a-condition  
https://www.marsja.se/r-add-column-to-dataframe-based-on-other-columns-conditions-dplyr/  
